Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-03-30 joint_inf.tex
using \mid and fixing emptyset thing

For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i | \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

With this split formulation, define the mappings from patterns to splits as described above
$$
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
% the following seemed not necessary and a little confusing (what are j and k?)
% so that, for
% $$
% \patternToSplit(\alignmentColumn_i)=\siteSplit_j, \ \ancestralToSplit(\ancestralStateColumn_i)=\ancestralSplit_k.
% $$
Given a site-pattern-valued random variable $H$ and an ancestral-state-valued random variable $H$, define random variables
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV), \ \ancestralSplitRV := \ancestralToSplit(\ancestralStateColumnRV).
$$
We can now write the $i$th factor of \eqref{eq:full_likelihood} as
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)=P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)
$$
for some $j$ and $k$.
Taking the maximum of the probability of observing a particular column $i$ yields
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k | \tau, t) &=
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j| \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k |\siteSplitRV=\siteSplit_j, \tau, t) \\
&= P(\siteSplitRV=\siteSplit_j| \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k |\siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
%EM Just noticed that eta doesn't have to be a real partition, because they don't have to be disjoint, right?
Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

@@

\paragraph{Example}
%EM My long term suggestion is for you to put this in example in the main text and the above in the appendix.
Consider a fixed, binary four-tip tree $\tau$.
The set of all possible character assignments can be described as
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\{\emptyset\}, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips at which a character is assigned the character $1$.
For example, $\{\emptyset\}$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
$$
\mathcal{P}(\siteSplitSet) = \{\{\emptyset\}, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\}.
$$
Had we not identified character complements, we would need to consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
but the symmetry of the binary character model allows us to focus just on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\{\emptyset\}, \{1\}, \{2\}, \{1,2\}\}.
$$
We cannot similarly identify complements in this case since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\{\emptyset\}|\siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\}|\siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
%EM I dropped the \ancestralSplitPartition(\siteSplit, \tau, t) notation above. You can revert it if you like, but I prefer the bold eta notation as an ordered list of sets.
A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; define $\ancestralSplitPartition(\siteSplit, \tau, t)$ as the partition of $\mathcal{P}(\ancestralSplitSet)$ into those most likely ancestral splits for that particular site split, topology and set of branch lengths.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for site split $\emptyset$ are
$$
P(\ancestralSplitRV=\{\emptyset\}|\siteSplitRV=\{\emptyset\}, \tau, t) = 
(1-p)^5,
$$
$$
%EM Seems to me that this just requires one character change so would be p(1-p)^4.
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^2(1-p)^3,
$$
$$
%EM Have you told us in this example what the labeled topology is?
%Erick stopped here.
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
Here, the partition into most likely ancestral states will proceed as
$$
P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p)|\siteSplitRV=\{\emptyset\}, \tau, t) = 
(1-p)^5
$$
where the partition is the single element $\ancestralSplitPartition_j(\tau, p)=\{\emptyset\}$.
For the site split $\{1,3\}$ we have
$$
P(\ancestralSplitRV=\{\emptyset\}|\siteSplitRV=\{1,3\}, \tau, t) = 
P(\ancestralSplitRV=\{1,2\}|\siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\}|\siteSplitRV=\{1,3\}, \tau, t) = 
P(\ancestralSplitRV=\{2\}|\siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2,
$$
where
$$
P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p)|\siteSplitRV=\{1,3\}, \tau, t) = 
p^2(1-p)^3
$$
with the most likely partition $\ancestralSplitPartition_j(\tau, p)=\{\{\emptyset\},\{1,2\}\}.$

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Define \eqref{eq:site_pattern_likelihood} as
$$
L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j | \tau, t) \cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
$$
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j | \tau, t) +
            \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j | \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j | \tau, t) + \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j | \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j | \tau, t),
$$
and the partial log likelihood
$$
\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j | \tau^*, t^*)\cdot\log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) | \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} is written as
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau',t_1)},\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau^*,t_2)},\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\widehat{\fullAncestralSplitPartitions(\tau',t_1)}\}$ and $\{\hat{t}_2,\widehat{\fullAncestralSplitPartitions(\tau^*,t_2)}\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.

@@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
\multicolumn{3}{c}{Classical}\\
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j|\tau_1,t^*)$&$P(\siteSplitRV=\siteSplit_j|\tau_2,t^*)$\\
    \hline
    $\{\emptyset\}$&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
    $\{1,3\}$      &$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
    \hline
\multicolumn{3}{c}{Modified}\\
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j|\tau_1,t)$&$P(\siteSplitRV=\siteSplit_j|\tau_2,t)$\\
    \hline
    $\{\emptyset\}$&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

@@

    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t)|\siteSplitRV=\siteSplit_j,\tau,t)$\\

@@

    $\{\emptyset\}$&
    $\{\emptyset\}$&

@@

    $\{\emptyset\}$&

@@

    $\{\{\emptyset\},\{1,2,3\}\}$&

@@

                    \{\emptyset\}\\

@@

    $\{\emptyset\}$       &$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)^2$\\

@@

                    \{\{\emptyset\},\{1,2\}\}\\

>>>>

For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

With this split formulation, define the mappings from patterns to splits as described above
$$
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
% the following seemed not necessary and a little confusing (what are j and k?)
% so that, for
% $$
% \patternToSplit(\alignmentColumn_i)=\siteSplit_j, \ \ancestralToSplit(\ancestralStateColumn_i)=\ancestralSplit_k.
% $$
Given a site-pattern-valued random variable $H$ and an ancestral-state-valued random variable $H$, define random variables
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV), \ \ancestralSplitRV := \ancestralToSplit(\ancestralStateColumnRV).
$$
We can now write the $i$th factor of \eqref{eq:full_likelihood} as
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)=P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)
$$
for some $j$ and $k$.
Taking the maximum of the probability of observing a particular column $i$ yields
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &=
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
&= P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
%EM Just noticed that eta doesn't have to be a real partition, because they don't have to be disjoint, right?
Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

@@

\paragraph{Example}
%EM My long term suggestion is for you to put this in example in the main text and the above in the appendix.
Consider a fixed, binary four-tip tree $\tau$.
The set of all possible character assignments can be described as
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips at which a character is assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
$$
\mathcal{P}(\siteSplitSet) = \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\}.
$$
Had we not identified character complements, we would need to consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
but the symmetry of the binary character model allows us to focus just on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
We cannot similarly identify complements in this case since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
%EM I dropped the \ancestralSplitPartition(\siteSplit, \tau, t) notation above. You can revert it if you like, but I prefer the bold eta notation as an ordered list of sets.
A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; define $\ancestralSplitPartition(\siteSplit, \tau, t)$ as the partition of $\mathcal{P}(\ancestralSplitSet)$ into those most likely ancestral splits for that particular site split, topology and set of branch lengths.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for site split $\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
%EM Seems to me that this just requires one character change so would be p(1-p)^4.
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^2(1-p)^3,
$$
$$
%EM Have you told us in this example what the labeled topology is?
%Erick stopped here.
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
Here, the partition into most likely ancestral states will proceed as
$$
P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p) \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5
$$
where the partition is the single element $\ancestralSplitPartition_j(\tau, p)=\emptyset$.
For the site split $\{1,3\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2,
$$
where
$$
P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p) \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3
$$
with the most likely partition $\ancestralSplitPartition_j(\tau, p)=\{\emptyset,\{1,2\}\}.$

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Define \eqref{eq:site_pattern_likelihood} as
$$
L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
$$
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} is written as
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau',t_1)},\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau^*,t_2)},\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\widehat{\fullAncestralSplitPartitions(\tau',t_1)}\}$ and $\{\hat{t}_2,\widehat{\fullAncestralSplitPartitions(\tau^*,t_2)}\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.

@@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
\multicolumn{3}{c}{Classical}\\
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t^*)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t^*)$\\
    \hline
    $\emptyset$    &$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
    $\{1,3\}$      &$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
    \hline
\multicolumn{3}{c}{Modified}\\
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$     &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

@@

    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\

@@

    $\emptyset$&
    $\emptyset$&

@@

    $\emptyset$&

@@

    $\{\emptyset,\{1,2,3\}\}$&

@@

                    \emptyset\\

@@

    $\emptyset$       &$\emptyset$&$(1+x)^2   (1+w)(1+y)^2$\\

@@

                    \{\emptyset,\{1,2\}\}\\
